---
title: "Exercises for 03 - Visualize"
format: html
editor: visual
---

## Setup Code Chunk

Run this to reload your *chem* dataframe and set up the additional data sets we will use in the exercises.

```{r setup}
## remove all user defined objects in your workspace
rm(list=ls())

library(tidyverse)  # Provides functions used throughout this session

chem <- read_csv("data/chem_data.csv", 
    col_types = cols(gender = col_factor(NULL),
                     pregnancy_status_at_exam = col_factor(NULL),
                     collect_dt = col_datetime(format = "%Y/%m/%d %H:%M:%S"), 
                     receive_dt = col_datetime(format = "%Y/%m/%d %H:%M:%S"), 
                     verify_dt = col_datetime(format = "%Y/%m/%d %H:%M:%S"), 
                     pat_type = col_factor(NULL),
                     payor_groups = col_factor(NULL),
                     test = col_factor(NULL),
                     units = col_factor(NULL))) # Read in data - note we determined data types up front
# the code below prepares a data set that contains only potassium values
k_data <- chem |> 
  filter(test == 'potassium')
# the code below prepares a data set that contains electrolytes commonly assayed in a metabolic panel
lyte_data <- chem |> 
  filter(test %in% c('sodium', 'potassium', 'chloride', 'bicarbonate'))
```

**Stop Here**

## Creating a ggplot

```{r}

```

Does the plot fit the image you were expecting? Why or why not?

**Stop Here**

## The ggplot Template and Aesthetics

Let's recreate the histogram of potassium results.

You will need to:

1.  Pick a data set: *k_data*

2.  Choose a geom function: `geom_histogram()`

3.  Map aesthetic properties of the graph to variables of the data frame (in this case, the *value* variable should be on the x axis).

Fill in the blanks and run the code to create the plot.

(Hint: if you get stuck, take a look at the slide titled "ggplot()"!)

```{r}
#k_data <- chem |> # uncomment these lines if you do not have k_data in your environment
#  filter(test == 'potassium')
ggplot(data = ______) +
  ______(mapping = aes(x = ______))
```

When you run this code, you will most likely have what looks like an error displayed above the graph:

\`stat_bin()\` using \`bins = 30\`. Pick better value with \`binwidth\`.

This is actually just a message from R, telling you that it's a bad idea to use a default bin width.

The *value* variable holds the quantitative value a given test in this data set. For this specific analyte, what would be an appropriate binwidth for *value* to see a distribution aligned to the likely resolution of the measurements? (Visually inspect the data if you are less familiar with the range of typical potassium values.)

Now fill in the blanks below to recreate the code from the previous chunk, with the *binwidth* argument passed to the *geom_histogram* function.

```{r}
ggplot(______) +
  ______(mapping = aes(______), binwidth = ______)
```

The effect of reducing the bin width (and increasing the number of bins) is that the data is now presented in much greater detail.

Let's next add some color!

Create a new code chunk below. Copy and paste the code from the previous code chunk and add an aesthetic mapping that maps the *fill* aesthetic to the *gender* variable. (Hint: all aesthetics must be defined \*inside\* the aes() function!)

**Stop Here**

## Geom Functions

Run the following code chunk:

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1)
```

Now try to figure out how you would modify the code above so it draws a frequency polygon instead of histogram. One resource is the ggplot cheatsheet that can be found at <https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf>

```{r}

```

Modify the previous code chunk so that the frequency polygon's line color is blue. (Hint: color names go in quotes!)

```{r}

```

What do you think the following code will do? Try to predict what you'll see.

Then run it.

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1) +
  geom_freqpoly(mapping = aes(x = value), binwidth = 0.1)
```

**Stop Here**

## Faceting

A grid of subplots can be produced using faceting functions, including `facet_grid()` which we demonstrate below. You can think about faceting like grouping the data frame - it is a convenient method to generate plots of subsets of data. The code chunk below takes our *k_data* dataframe and generates a series of histograms that are split into 2 rows by *gender* and 6 columns based on the *pat_type*, so that each individual plot represents the distribution of the intersection of those 2 variables.

```{r}
ggplot(data = k_data) +
  geom_histogram(mapping = aes(x = value), binwidth = 0.1) +
  facet_grid(rows = vars(gender), 
		   cols = vars(pat_type),
		   scales = "free")
```

Recall that in our setup chunk we made a data frame that contains all of the electrolyte data *lyte_data*. We would like to visualize the distribution of results in this data set by test and by patient gender. Build a faceted plot for the data frame that separates each gender group into separate rows and each test into separate columns. Use the `scales = "free"` argument to allow each column to use a different scale.

```{r}
#lyte_data <- chem |> # uncomment these lines if you do not have lyte_data in your environment
#  filter(test %in% c('sodium', 'potassium', 'bicarbonate', 'chloride'))
ggplot(data = ......) +
  geom_histogram(mapping = aes(x = ......)) +
  facet_grid(cols = vars(......), 
             rows = vars(......), 
             .......)
```

The histogram shapes have some irregularities. Why might that be? How can you modify the plot to provide a better visualization? Consider that each of the analytes has a different measurement range. What option(s) for your geom function can you use to handle these disparate values?

```{r}
ggplot(data = ......) +
  geom_histogram(mapping = aes(x = ......), .......) +
  facet_grid(cols = vars(......), 
             rows = vars(......), 
             .......)
```

**Stop Here**
