---
title: "Grouping and Summarizing"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Setup

```{r setup}
library(tidyverse)  # Provides functions used throughout this session

covid_testing <- read_csv("covid_testing.csv")
```

## Summarize

### How many tests are performed per day?
```{r}
ggplot(data = covid_testing) +
  geom_histogram(aes(x = pan_day), binwidth = 1)
```

### Row counting functions

There are many functions built to work with `summarize()` that can help with counting rows. `n()` returns the number of rows and `n_distinct()` returns the number of distinct (unique) rows for a given variable.

```{r}
covid_testing |> 
  select(mrn, pan_day) |> 
  head(4) |> 
	summarize(order_count = n(),
	          day_count    = n_distinct(pan_day)) 
```

### Calculations within summarize

Within the `summarize()` function, you can perform calculations on the new variables you created.

```{r, eval=FALSE}
covid_testing |> 
	summarize(order_count = n(),
	          day_count = n_distinct(pan_day),
	          orders_per_day = order_count/day_count)
```

## Your Turn 1

Fill in gaps in code below to calculate the average count of orders per clinic

```{r, eval=FALSE}
stats_1b <- covid_testing |>
	summarize(order_count = n(),
	          clinic_count = ______________),
	          orders_per_clinic = _________________)
stats_1b
```

## Summarize examples

### Last pandemic day (in data)

```{r, eval=FALSE}
covid_testing |>
  summarize(last_day = ____(pan_day))
```

### Median turnaround time

```{r, eval=FALSE}
covid_testing |>
  mutate(col_ver_tat = col_rec_tat + rec_ver_tat) |>
  summarize(col_ver_tat_mean = mean(col_ver_tat))
```

### Median and 75th percentile turnaround time

```{r, eval = FALSE}
covid_testing |>
  mutate(col_ver_tat = col_rec_tat + rec_ver_tat) |>
  summarize(col_ver_tat_mean = mean(col_ver_tat),
		    col_ver_75_pctile = quantile(col_ver_tat, 0.75)
```

## Your Turn 2

For the covid_testing data frame, calculate both the median and the 95th percentile collect-to-verify turnaround time.

```{r}
covid_testing |> 
  mutate(col_ver_tat = col_rec_tat + rec_ver_tat) |> 
  summarize(___________ = ______________,
            ___________ = ______________)
```

## group_by

```{r}
covid_testing |> 
  group_by(pan_day)
```

## group_by |> summarize

```{r}
covid_testing |>
	group_by(pan_day) |>
	summarize(order_count = n())
```

## Your Turn 3

Calculate: 
a) The median turnaround time for each day

```{r, eval=FALSE}
covid_testing  |> 
  mutate(col_ver_tat = ________________) |> 
  group_by(______) |> 
  summarize(col_ver_tat_median = ___________)
```

b) The median collect-to-verify turnaround time for each clinic/unit

```{r, eval=FALSE}
covid_testing |> 
  mutate(col_ver_tat = ________________) |> 
  group_by(_________) |> 
  summarize(col_ver_tat_median = ___________)
```

c) The median number of orders per day

```{r}
covid_testing |> 
   group_by(_________) |> 
   summarize(_________) |> 
  
```

## group_by |> summarize examples

### 95th percentile turnaround time per clinic

```{r}
covid_testing |>
  group_by(clinic_name) |>
  summarize(tat_95th_pctile = quantile(rec_ver_tat, 0.95))
```

### Number of tests per day
```{r, eval=FALSE}
tmp <- covid_testing |>
  group_by(pan_day) |>
  summarize(n_tests = n())

ggplot(data = tmp) +
  geom_point(aes(x = pan_day, y = n_tests)) +
  geom_smooth(aes(x = pan_day, y = n_tests), method = 'loess') +
  ylab("# of tests") +
  xlab("Pandemic day")
```

## What else

### Test for association
```{r}
covid_testing_2 <- covid_testing |>
  mutate(payor_group_fac = case_when(
				is.na(payor_group) 			     ~ "unassigned",
				payor_group %in% c("charity care", 
						    "medical assistance", 
						    "self pay", 
					            "other") 		     ~ "other",
									TRUE ~ payor_group)
  ) |>
  filter(result %in% c("positive", "negative"))
```

```{r}
# Generate counts
tmp_table_tall <- covid_testing_2 |> 
  group_by(payor_group_fac, result) |>
  summarize(n = n()) |>
  ungroup()
tmp_table_tall

# Pivot from tall to wide table
tmp_table_wide <- tmp_table_tall |>
  spread(key = "result", value = "n")
tmp_table_wide
```

```{r}
# Hypergeometric test for association
tmp_table_wide |>
  select(-payor_group_fac) |>
  fisher.test(simulate.p.value = T)
```

### Logistic Regression
```{r}
tmp <- covid_testing_2 |>
  filter(payor_group_fac %in% c("commercial", "government")) |>
  mutate(result_fac = factor(result, 
                             levels=c("negative", "positive"), 
                             ordered=T),
         payor_group_fac = (payor_group == "government"))
tmp_fit <- glm(result_fac ~ payor_group_fac + age,  	# model formula
               data = tmp,                           	# dataset
               family = "binomial"                    # type of model
  )
summary(tmp_fit)
exp(coefficients(tmp_fit))  					                # odds
```

