---
title: "Exercise for 01 - Fixing Syntax Mistakes"
output: html_document
---

I’ve written some code to create a sequence from 0 to 500 in increments of 25 called sequence_of_25s.  Ultimately I want to calculate the median value of this sequence.  Unfortunately I’ve made some mistakes in my code and I am hoping you can help me find them.

```{r}

sequence_of_25s -< seq(from=0 to=50, by=25)

```

My first code chunk (above) includes 3 errors:
1 - The assignment operator is typed incorrectly
2 - I am missing a comma to separate my "from" and "to" arguments
3 - I intended to create a sequence from 0-500, but I accidentally typed 50 instead

Below is the corrected code

```{r}

sequence_of_25s <- seq(from=0, to=500, by=25)

```

```{r}

median(sequence of_25s]

```

My second code chunk chunk (above) includes 2 errors:
1 - I misspelled my object name sequence_of_25s
2 - I closed my median function with a square bracket "]" instead of a parenthesis ")"

Below is the corrected code

```{r}

median(sequence_of_25s)

```

