---
title: "Transform Data"
---

## Reload data set

```{r, message=FALSE}
library(tidyverse)  # Provides functions used throughout this session

covid_testing <- read_csv("covid_testing.csv") #Read in data
```


# Arrange

```{r arrange, eval=FALSE}
arrange(covid_testing, first_name) #Sort in alphabetical order
arrange(covid_testing, desc(mrn))  #Sort from largest to smallest MRN
```

## Your Turn 1

The column `ct_value` contains the cycle threshold (Ct) for the real-time PCR that generated the final result.

How might you use arrange() to determine the highest and lowest Ct result in the dataset? (Hint: Try sorting the rows by the Ct result value contained in the column named `ct_value`) 

```{r Your Turn 1, eval=FALSE}
#replace the ____ with the name of the column you would like to arrange the dataframe by.
arrange(covid_testing, ct_value)
arrange(covid_testing, desc(ct_value))
```


## Pop Quiz

```{r pop quiz}
toy_data <- data.frame(nums=c(50,1000,3),chars=c("50","1000","3"))
arrange(toy_data,nums)
arrange(toy_data,chars)
```

## Steps and the pipe

```{r Steps and the pipe}
# These 3 lines of code are an example of the how we would accomplish a multi-step analysis based on what we've learned so far. 
day_10 <- filter(covid_testing, pan_day <= 10)
day_10 <- select(day_10, clinic_name)
day_10 <- arrange(day_10 , clinic_name)

# vs

# This code accomplishes the same task and is far more legible. 
covid_testing |> 
  filter(pan_day <= 10) |> 
  select(clinic_name) |> 
  arrange(clinic_name)
```

## Your Turn 2

Use `|>` in the following chunk to write a sequence of functions that: 

1. Filters to tests from the clinic (`clinic_name`) of "picu"

2. Selects the column with the receive to verify turnaround time (`rec_ver_tat`) as well as the day from start of the pandemic (`pan_day`)

3. Arrange the `pan_day` from highest to lowest so you can see most recent tests on top.

Using "<-", assign the result to a new variable, call it whatever you want.

```{r Your Turn 2, eval=FALSE}
covid_testing |>  
  filter(clinic_name == "picu") |>  #Here filter to the clinic with the name "picu"
  select(pan_day, rec_ver_tat) |>  #Here select the columns you want to keep
  arrange(desc(pan_day))    #Here arrange the pan_day column from highest to lowest (HINT: See the arrange section of this file to see how to use the desc() function)
```


## Pipe Bonus Challenge (only if you have time)

Use your mouse to select the name of the new data frame from the list of objects in the Environment tab in the upper-right pane of Rstudio. Answer the following questions:

- Can you find the test that exceeded 10 hours in TAT?
- What is the typical TAT for the PICU?
- Given this data, does it appear that there is a chronic problem in TAT for the PICU, or was this test an outlier?


# mutate

```{r}
# The following code creates a data frame with a new column called c_r_tat_mins which contains the collect to receive TAT quantified in minutes.
covid_testing |>
	mutate(c_r_tat_mins = col_rec_tat*60)
```

## Your Turn 3

Create a new column using the mutate() function that contains the total TAT. 
Hint: The total TAT is the sum of `col_rec_tat` and `rec_ver_tat`

```{r Your Turn 3, eval=FALSE}
covid_testing  |> 
	mutate(total_tat = col_rec_tat + rec_ver_tat)
```

## Replace a column with `mutate()`

```{r}
covid_testing |> 
	mutate(mrn = as.character(mrn))
```

